---
title: "Pilpres 2024 — Bayesian Polling from TPS Data"
author: "Derrian Gharda"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

## 1 Setup

```{r}
#| label: setup

library(tidyverse)
library(scales)
library(knitr)
library(jsonlite)
library(ggdist)
```

## 2 Load & Clean Data

```{r}
#| label: load-data
#| cache: true

kode_provinsi <- c(
  "11" = "ACEH",
  "12" = "SUMATERA UTARA",
  "13" = "SUMATERA BARAT",
  "14" = "RIAU",
  "15" = "JAMBI",
  "16" = "SUMATERA SELATAN",
  "17" = "BENGKULU",
  "18" = "LAMPUNG",
  "19" = "KEPULAUAN BANGKA BELITUNG",
  "21" = "KEPULAUAN RIAU",
  "31" = "DKI JAKARTA",
  "32" = "JAWA BARAT",
  "33" = "JAWA TENGAH",
  "34" = "DI YOGYAKARTA",
  "35" = "JAWA TIMUR",
  "36" = "BANTEN",
  "51" = "BALI",
  "52" = "NUSA TENGGARA BARAT",
  "53" = "NUSA TENGGARA TIMUR",
  "61" = "KALIMANTAN BARAT",
  "62" = "KALIMANTAN TENGAH",
  "63" = "KALIMANTAN SELATAN",
  "64" = "KALIMANTAN TIMUR",
  "65" = "KALIMANTAN UTARA",
  "71" = "SULAWESI UTARA",
  "72" = "SULAWESI TENGAH",
  "73" = "SULAWESI SELATAN",
  "74" = "SULAWESI TENGGARA",
  "75" = "GORONTALO",
  "76" = "SULAWESI BARAT",
  "81" = "MALUKU",
  "82" = "MALUKU UTARA",
  "91" = "PAPUA BARAT DAYA",
  "92" = "PAPUA",
  "93" = "PAPUA SELATAN",
  "94" = "PAPUA TENGAH",
  "95" = "PAPUA PEGUNUNGAN",
  "96" = "PAPUA BARAT"
)

tps_raw <- read_csv("data/tps_raw.csv", col_types = cols(tps = col_character()))

tps <- tps_raw |>
  mutate(
    kode_prov = str_sub(tps, 1, 2),
    provinsi  = kode_provinsi[kode_prov],
    suara_01  = paslon01,
    suara_02  = paslon02,
    suara_03  = paslon03,
    p_anies   = suara_01 / total_suara,
    p_prabowo = suara_02 / total_suara,
    p_ganjar  = suara_03 / total_suara
  ) |>
  filter(total_suara > 0, !is.na(provinsi))
```

## 3 TPS Allocation by Province Size Cluster

Before sampling, the polling agency groups all 38 provinces into four tiers by DPT weight
and assigns a fixed TPS visit count to every province within each tier. Larger tiers receive
more visits than smaller ones, but every province inside the same tier gets exactly the same
count — regardless of how much their voter populations differ within the tier.

This deliberate imperfection means a province's share of the raw sample will not match its
true DPT share. The gap is most visible inside the largest tier, where the biggest and
smallest member receive identical TPS counts despite very different voter populations.
Section 5 visualises this gap; DPT weighting in Section 9 corrects it.

```{r}
#| label: province-allocation

province_dpt <- tps |>
  group_by(provinsi) |>
  summarise(total_voters = sum(total_suara)) |>
  mutate(proportion = total_voters / sum(total_voters))

cluster_tps <- c("1" = 10L, "2" = 30L, "3" = 60L, "4" = 100L)

province_allocation <- province_dpt |>
  mutate(
    cluster     = as.character(ntile(proportion, 4)),
    n_allocated = cluster_tps[cluster]
  )

cat("Total TPS allocated:", sum(province_allocation$n_allocated), "\n")

province_allocation |>
  arrange(desc(n_allocated), desc(proportion)) |>
  mutate(
    total_voters = scales::comma(total_voters),
    proportion   = percent(proportion, accuracy = 0.01),
    cluster      = paste0("Tier ", cluster, " (", cluster_tps[cluster], " TPS)")
  ) |>
  select(provinsi, total_voters, proportion, cluster, n_allocated) |>
  knitr::kable(col.names = c("Province", "Registered Voters", "DPT Weight", "Tier", "Allocated TPS"))
```

## 4 Polling Sample — Cluster-Based TPS per Province

With the per-province allocation defined, the agency draws the assigned number of TPS at
random from each province's pool.

```{r}
#| label: stratified-sample

set.seed(42)
stratified_sample <- tps |>
  group_by(provinsi) |>
  group_modify(function(.x, .y) {
    n <- province_allocation$n_allocated[province_allocation$provinsi == .y$provinsi]
    slice_sample(.x, n = min(n, nrow(.x)))
  }) |>
  ungroup()

cat("Total TPS sampled:", nrow(stratified_sample), "\n")
cat("Total TPS in dataset:", nrow(tps), "\n")
cat("Coverage:", percent(nrow(stratified_sample) / nrow(tps), accuracy = 0.01), "of all TPS\n")
```

## 5 Why Weighting Is Necessary

Within each tier, every province contributes the same number of TPS to the raw sample —
so a province with twice the voters of its tier-mate contributes equally before weighting
but should contribute twice as much to the national estimate. DPT weighting replaces each
province's raw sample share with its true voter share.

The slope chart shows how contributions shift. Large provinces in upper tiers (lines going
steeply **up**) are heavily under-represented in the raw sample; smaller provinces in those
same tiers (lines going **down**) are over-represented. The effect is far more visible here
than it would be under a proportional allocation.

```{r}
#| label: fig-stratification-check
#| fig-cap: "Each province's raw sample share (left, equal within tier) vs its true DPT weight (right) — cluster-based allocation creates the large gap DPT weighting corrects"
#| fig-height: 7

sample_props <- stratified_sample |>
  count(provinsi, name = "n_sampled") |>
  mutate(sample_prop = n_sampled / sum(n_sampled))

highlight <- province_dpt |>
  arrange(desc(proportion)) |>
  slice(c(1:4, (n() - 3):n())) |>
  pull(provinsi)

province_dpt |>
  left_join(sample_props, by = "provinsi") |>
  pivot_longer(c(sample_prop, proportion),
               names_to = "stage", values_to = "contribution") |>
  mutate(stage = factor(stage,
    levels = c("sample_prop", "proportion"),
    labels = c("Before\n(sample proportion)", "After\n(DPT-weighted)")
  ),
  is_highlight = provinsi %in% highlight) |>
  ggplot(aes(x = stage, y = contribution, group = provinsi)) +
  geom_line(aes(color = is_highlight, alpha = is_highlight, linewidth = is_highlight)) +
  geom_point(aes(color = is_highlight), size = 2) +
  geom_text(
    data = ~ filter(.x, stage == "After\n(DPT-weighted)", is_highlight),
    aes(label = provinsi),
    hjust = -0.08, size = 3, color = "#2c7bb6"
  ) +
  scale_y_continuous(labels = percent_format()) +
  scale_color_manual(values = c("gray75", "#2c7bb6"), guide = "none") +
  scale_alpha_manual(values = c(0.4, 1), guide = "none") +
  scale_linewidth_manual(values = c(0.5, 1.2), guide = "none") +
  labs(
    title = "Province Contribution: Before vs After DPT Weighting",
    subtitle = "Highlighted: 4 largest and 4 smallest provinces by registered voters",
    x = NULL, y = "Contribution to national estimate"
  ) +
  theme_minimal(base_size = 13) +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(r = 140))
```

## 6 Bayesian Beta-Binomial Model

Prior: Beta(1,1) — flat, no prior knowledge.
Posterior: Beta(1 + k, 1 + n − k).
Run independently for each of the three candidates.

```{r}
#| label: bayesian-model

province_votes_long <- stratified_sample |>
  group_by(provinsi) |>
  summarise(
    k_01 = sum(suara_01),
    k_02 = sum(suara_02),
    k_03 = sum(suara_03),
    n    = sum(total_suara),
    .groups = "drop"
  ) |>
  pivot_longer(c(k_01, k_02, k_03), names_to = "candidate", values_to = "k") |>
  mutate(candidate = recode(candidate,
    k_01 = "Anies (01)",
    k_02 = "Prabowo (02)",
    k_03 = "Ganjar (03)"
  ))

province_bayes <- province_votes_long |>
  mutate(
    alpha_post = 1 + k,
    beta_post  = 1 + (n - k),
    mean_post  = alpha_post / (alpha_post + beta_post),
    ci_lower   = qbeta(0.025, alpha_post, beta_post),
    ci_upper   = qbeta(0.975, alpha_post, beta_post)
  )

sample_provs <- c("DKI JAKARTA", "SUMATERA UTARA")

province_bayes |>
  filter(provinsi %in% sample_provs) |>
  arrange(provinsi, candidate) |>
  select(provinsi, candidate, k, n, mean_post, ci_lower, ci_upper) |>
  mutate(across(c(mean_post, ci_lower, ci_upper), ~ percent(.x, accuracy = 0.1))) |>
  knitr::kable(col.names = c("Provinsi", "Candidate", "Votes", "Total Votes",
                              "Estimate", "CI Lower", "CI Upper"))
```

## 7 Province Estimates — Posterior Distributions

Two provinces chosen to contrast outcomes: **SUMATERA UTARA** (estimate close to truth) and
**DKI JAKARTA** (estimate noticeably off). Even with proportionally allocated TPS the sample
can be unlucky — polling stations drawn from a large, diverse city like Jakarta are more likely
to hit a non-representative pocket than in a more homogeneous province. Whether this matters
nationally depends on each province's share of registered voters (DPT), shown in the row labels.

```{r}
#| label: fig-province-estimates
#| fig-cap: "Blue = Beta posterior from proportionally allocated TPS · Red line = true province result · Top-left annotation = point estimate, truth, error"
#| fig-height: 6
#| fig-width: 11

province_actual <- tps |>
  group_by(provinsi) |>
  summarise(
    actual_01 = sum(suara_01) / sum(total_suara),
    actual_02 = sum(suara_02) / sum(total_suara),
    actual_03 = sum(suara_03) / sum(total_suara),
    .groups = "drop"
  ) |>
  pivot_longer(c(actual_01, actual_02, actual_03),
               names_to = "candidate", values_to = "actual") |>
  mutate(candidate = recode(candidate,
    actual_01 = "Anies (01)",
    actual_02 = "Prabowo (02)",
    actual_03 = "Ganjar (03)"
  ))

panel_ann <- province_bayes |>
  filter(provinsi %in% sample_provs) |>
  left_join(province_actual, by = c("provinsi", "candidate")) |>
  left_join(province_dpt |> select(provinsi, proportion), by = "provinsi") |>
  mutate(
    error_pp = (mean_post - actual) * 100,
    ann = paste0(
      "Est: ",   percent(mean_post, accuracy = 0.1), "\n",
      "True: ",  percent(actual,    accuracy = 0.1), "\n",
      "Error: ", sprintf("%+.1f pp", error_pp)
    )
  )

prov_labels <- province_dpt |>
  filter(provinsi %in% sample_provs) |>
  mutate(label = paste0(provinsi, "  (DPT: ", percent(proportion, accuracy = 0.1), ")")) |>
  select(provinsi, label) |>
  deframe()

dist_data <- panel_ann |>
  rowwise() |>
  mutate(
    x_lo  = max(0, min(ci_lower, actual) - 0.05),
    x_hi  = min(1, max(ci_upper, actual) + 0.05),
    curve = list(tibble(
      x = seq(x_lo, x_hi, length.out = 600),
      y = dbeta(seq(x_lo, x_hi, length.out = 600), alpha_post, beta_post)
    ))
  ) |>
  select(-x_lo, -x_hi) |>
  unnest(curve)

ggplot(dist_data, aes(x = x, y = y)) +
  geom_area(fill = "#2c7bb6", alpha = 0.3) +
  geom_line(color = "#2c7bb6", linewidth = 0.9) +
  geom_vline(aes(xintercept = actual), color = "#d73027", linewidth = 1.2) +
  geom_text(
    data = panel_ann,
    aes(x = -Inf, y = Inf, label = ann),
    hjust = -0.08, vjust = 1.25, size = 3.2, color = "gray30",
    inherit.aes = FALSE
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  facet_grid(provinsi ~ candidate, scales = "free",
             labeller = labeller(provinsi = prov_labels)) +
  labs(
    title = "Posterior Distributions vs True Province Result",
    x = "Vote share", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    strip.text.y       = element_text(angle = 0, hjust = 0, face = "bold", size = 11),
    strip.text.x       = element_text(face = "bold", size = 12)
  )
```

## 8 What Happens When We Increase n for DKI Jakarta?

Section 7 showed that DKI Jakarta's estimate was noticeably off. This section asks: what if the
polling agency had sent more surveyors to Jakarta? We start from Jakarta's cluster allocation of
60 TPS and scale up to 600 and 6000. Each larger set is a strict superset of the previous —
the same 60 TPS, plus 540 more, plus 5940 more — so the distributions show genuine learning
rather than different random luck.

```{r}
#| label: jakarta-sample-size

jakarta_n60_tps <- stratified_sample |> filter(provinsi == "DKI JAKARTA")

set.seed(99)
jakarta_extra <- tps |>
  filter(provinsi == "DKI JAKARTA") |>
  anti_join(jakarta_n60_tps |> select(tps), by = "tps") |>
  slice_sample(n = 5940)

compute_posterior <- function(tps_data, n_label) {
  tps_data |>
    summarise(
      k_01 = sum(suara_01), k_02 = sum(suara_02), k_03 = sum(suara_03),
      n    = sum(total_suara)
    ) |>
    pivot_longer(c(k_01, k_02, k_03), names_to = "candidate", values_to = "k") |>
    mutate(
      candidate  = recode(candidate,
        k_01 = "Anies (01)", k_02 = "Prabowo (02)", k_03 = "Ganjar (03)"),
      n_label    = n_label,
      alpha_post = 1 + k,
      beta_post  = 1 + (n - k),
      mean_post  = alpha_post / (alpha_post + beta_post),
      ci_lower   = qbeta(0.025, alpha_post, beta_post),
      ci_upper   = qbeta(0.975, alpha_post, beta_post)
    )
}

jakarta_compare <- bind_rows(
  compute_posterior(jakarta_n60_tps,                                               "n = 60 TPS"),
  compute_posterior(bind_rows(jakarta_n60_tps, slice_head(jakarta_extra, n=540)),  "n = 600 TPS"),
  compute_posterior(bind_rows(jakarta_n60_tps, jakarta_extra),                     "n = 6000 TPS")
) |>
  mutate(n_label = factor(n_label, levels = c("n = 60 TPS", "n = 600 TPS", "n = 6000 TPS")))

jakarta_actual <- province_actual |> filter(provinsi == "DKI JAKARTA")
```

```{r}
#| label: fig-jakarta-sample-size
#| fig-cap: "Each distribution is built on the same TPS as the one before it, plus more. Red line = true DKI Jakarta result."
#| fig-height: 5

n_colors <- c("n = 60 TPS"   = "#a8c8e8",
               "n = 600 TPS"  = "#3d8ecf",
               "n = 6000 TPS" = "#1a3d6e")

x_bounds <- jakarta_compare |>
  left_join(jakarta_actual, by = "candidate") |>
  group_by(candidate) |>
  summarise(
    x_lo = max(0, min(ci_lower, actual) - 0.04),
    x_hi = min(1, max(ci_upper, actual) + 0.04),
    .groups = "drop"
  )

dist_jakarta <- jakarta_compare |>
  left_join(jakarta_actual,  by = "candidate") |>
  left_join(x_bounds,        by = "candidate") |>
  rowwise() |>
  mutate(curve = list(tibble(
    x = seq(x_lo, x_hi, length.out = 600),
    y = dbeta(seq(x_lo, x_hi, length.out = 600), alpha_post, beta_post)
  ))) |>
  select(-x_lo, -x_hi) |>
  unnest(curve)

jakarta_ann <- jakarta_compare |>
  left_join(jakarta_actual, by = "candidate") |>
  mutate(
    error_pp = (mean_post - actual) * 100,
    x_ann    = (alpha_post - 1) / (alpha_post + beta_post - 2),
    y_ann    = dbeta(x_ann, alpha_post, beta_post),
    label    = sprintf("%+.1f pp", error_pp)
  )

ggplot(dist_jakarta, aes(x = x, y = y, color = n_label, fill = n_label)) +
  geom_area(alpha = 0.18, position = "identity") +
  geom_line(linewidth = 0.9) +
  geom_vline(
    data = jakarta_actual, aes(xintercept = actual),
    color = "#d73027", linewidth = 1.2, inherit.aes = FALSE
  ) +
  geom_text(
    data = jakarta_ann,
    aes(x = x_ann, y = y_ann, label = label, color = n_label),
    vjust = -0.4, size = 3, fontface = "bold",
    inherit.aes = FALSE, show.legend = FALSE
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = n_colors) +
  scale_fill_manual(values  = n_colors) +
  facet_wrap(~ candidate, scales = "free_x") +
  labs(
    title = "DKI Jakarta — Effect of Increasing TPS Sample Size",
    subtitle = "Larger n → narrower posterior, closer to the red truth line",
    x = "Vote share", y = NULL, color = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position    = "top",
    strip.text         = element_text(face = "bold")
  )
```

```{r}
#| label: jakarta-accuracy-table

jakarta_compare |>
  left_join(jakarta_actual, by = "candidate") |>
  mutate(
    error_pp = (mean_post - actual) * 100,
    cell     = paste0(percent(mean_post, accuracy = 0.1),
                      " (", sprintf("%+.1f pp", error_pp), ")")
  ) |>
  select(candidate, n_label, cell) |>
  pivot_wider(names_from = n_label, values_from = cell) |>
  left_join(
    jakarta_actual |> mutate(true = percent(actual, accuracy = 0.1)) |> select(candidate, true),
    by = "candidate"
  ) |>
  select(candidate, `n = 60 TPS`, `n = 600 TPS`, `n = 6000 TPS`, true) |>
  knitr::kable(col.names = c("Candidate",
                              "Estimate & Error (n = 60)",
                              "Estimate & Error (n = 600)",
                              "Estimate & Error (n = 6000)",
                              "True"))
```

## 9 National Weighted Estimate vs True Result

Before computing the final national estimate, it is worth asking: do the provincial errors we
observed actually matter at the national level? The impact of any single province's miss is
bounded by its share of registered voters — a large error in a small province is heavily diluted.

```{r}
#| label: national-impact

panel_ann |>
  mutate(national_impact_pp = abs(error_pp) * proportion) |>
  select(provinsi, candidate, error_pp, proportion, national_impact_pp) |>
  mutate(
    error_pp           = sprintf("%+.2f pp", error_pp),
    proportion         = percent(proportion, accuracy = 0.1),
    national_impact_pp = sprintf("%.3f pp", national_impact_pp)
  ) |>
  knitr::kable(col.names = c("Province", "Candidate",
                              "Province error", "DPT weight",
                              "National impact (error × weight)"))
```

With that context established, we now aggregate all province posteriors into a national estimate.
Province-level posteriors are weighted by each province's DPT share, correcting for the unequal
proportional allocation in Section 4. A naive (unweighted) aggregate treats every province as equal
and is shown alongside as a baseline.

```{r}
#| label: national-estimate

actuals <- tibble(
  candidate = c("Anies (01)", "Prabowo (02)", "Ganjar (03)"),
  actual    = c(0.2495, 0.5859, 0.1647)
)

national_est <- province_bayes |>
  left_join(province_dpt |> select(provinsi, proportion), by = "provinsi") |>
  group_by(candidate) |>
  summarise(
    naive_mean     = mean(mean_post),
    weighted_mean  = sum(mean_post  * proportion),
    weighted_lower = sum(ci_lower   * proportion),
    weighted_upper = sum(ci_upper   * proportion),
    .groups = "drop"
  ) |>
  left_join(actuals, by = "candidate") |>
  mutate(
    naive_error    = naive_mean    - actual,
    weighted_error = weighted_mean - actual
  )

national_est |>
  mutate(across(c(naive_mean, weighted_mean, weighted_lower, weighted_upper, actual),
                ~ percent(.x, accuracy = 0.01)),
         across(c(naive_error, weighted_error), ~ percent(.x, accuracy = 0.01))) |>
  knitr::kable(col.names = c("Candidate", "Naive", "Weighted", "CI Lower", "CI Upper",
                              "True", "Naive Error", "Weighted Error"))
```

```{r}
#| label: fig-national-estimate
#| fig-cap: "Weighted estimate (blue) closely tracks the true result (red). Naive estimate (gray) is biased by over-weighting small provinces."

national_est |>
  pivot_longer(c(naive_mean, weighted_mean, actual),
               names_to = "type", values_to = "value") |>
  mutate(type = recode(type,
    naive_mean    = "Naive (unweighted)",
    weighted_mean = "DPT-weighted estimate",
    actual        = "True result"
  ),
  type = factor(type, levels = c("True result", "DPT-weighted estimate", "Naive (unweighted)"))) |>
  ggplot(aes(x = candidate, y = value, color = type, shape = type)) +
  geom_errorbar(
    data = national_est,
    aes(x = candidate, ymin = weighted_lower, ymax = weighted_upper),
    inherit.aes = FALSE,
    width = 0.15, color = "#2c7bb6", linewidth = 0.8
  ) +
  geom_point(size = 4, position = position_dodge(width = 0.4)) +
  scale_y_continuous(labels = percent_format()) +
  scale_color_manual(values = c("True result" = "#d73027",
                                "DPT-weighted estimate" = "#2c7bb6",
                                "Naive (unweighted)" = "gray50")) +
  scale_shape_manual(values = c("True result" = 18,
                                "DPT-weighted estimate" = 16,
                                "Naive (unweighted)" = 15)) +
  labs(
    title = "National Estimate: Weighted vs Naive vs True",
    subtitle = paste0("Polling agency sampled ", nrow(stratified_sample),
                      " TPS proportionally allocated by province (",
                      percent(nrow(stratified_sample) / nrow(tps), accuracy = 0.01),
                      " of all TPS)"),
    x = NULL, y = "Estimated national support", color = NULL, shape = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

